Syntax and Semantics of Cedille

### 2021-04-13 docs/semantics/paper.tex
docs: semantics: typo
docs: semantics: make delta subterm synthesizing

\newcommand{\choice}[0]{\zeta}
\newcommand{\elcap}[0]{\cap}
\newcommand{\abs}[4]{{#1}\, #2\! : \! #3.\, #4}
\newcommand{\absu}[3]{{#1}\, #2.\, #3}
\mathchardef\mhyph="2D % Define a "math hyphen"
\newcommand{\interp}[1]{\llbracket #1 \rrbracket} 
\newcommand{\tpcheck}[0]{\Leftarrow}
\newcommand{\tpsynth}[0]{\Rightarrow}
\newcommand{\tpsynthleads}[0]{\ensuremath{\overset{\leadsto_\beta^*}{\Rightarrow}}}
\newcommand{\cbe}[0]{c\beta\eta}
\newcommand{\utp}[0]{\mathcal{U}}
\newcommand{\startcase}[1]{\vspace{#1} \noindent\textbf{\underline{Case:}}}
\newcommand{\startsubcase}[1]{\vspace{#1} \noindent\textbf{\underline{Subcase:}}}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{observation}[theorem]{Observation}

@@

In terms, when the type of an introduction form is checked or the type of an
elimination form is synthesized, we use call-by-name normalization for types
\(\leadsto_\beta^*\) to put types in weak head normal form, revealing type
constructors.
We abbreviate the conjuction ``term \(t\) synthesizes some type, and that type
call-by-name reduces to another type'' with the symbol \(\tpsynthleads\),
defined formally at the top of Figure~\ref{fig:tp}.
When a redex (reducible expression) occurs such that the argument to a type is a
term, such as \((\abs{\lambda}{x}{T_1}{T_2})\ t_1\), the reduction uses
substitution with annotations to \([t_1/x]^{T_1}\ T_2\).

@@

\begin{figure}
  \[
    \begin{array}{c}
      \Gamma \vdash t \tpsynthleads T\ =\ \exists T'.\ \Gamma
      \vdash t \tpsynth T' \land T' \leadsto_\beta^* T
      \\ \\ \\
  \begin{array}{cc}
    \infer{\Gamma\vdash x\tpsynth T}{(x : T)\in\Gamma} &
    \infer{\Gamma\vdash t\tpcheck T}{\Gamma\vdash t\tpsynth T' & T' \cong T} \\ \\
    \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck T}{T \leadsto_\beta^* \abs{\Pi}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t\tpcheck T_2} &
    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynthleads \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\

    \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck T'}
          {T' \leadsto_\beta^* \abs{\forall}{X}{\kappa}{T} & \Gamma,X:\kappa\vdash t \tpcheck T} &
    \infer{\Gamma\vdash t \cdot T' \tpsynth [T'/X]T}
          {\Gamma\vdash t \tpsynthleads \abs{\forall}{X}{\kappa}{T} & \Gamma\vdash T' \tpsynth \kappa' & \kappa'\cong\kappa} \\ \\

    \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck T}
          {T \leadsto_\beta^* \abs{\forall}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t \tpcheck T_2 & x\not\in\textit{FV}(|t|)} &
    \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynthleads \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\

    \infer{\Gamma\vdash [ t_1 , t_2 ] \tpcheck T}
          {
           \begin{array}{cc}
             T \leadsto_\beta^* \abs{\iota}{x}{T_1}{T_1}
             & \Gamma \vdash t_1 \tpcheck T_1
             \\ \Gamma \vdash t_2 \tpcheck [t_1/x]^{T_1}\ T_2
             & |t_1| =_{\beta\eta} |t_2|
           \end{array}
          } &
    \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynthleads \abs{\iota}{x}{T}{T'}} \\ \\
    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x] T'}{\Gamma\vdash t \tpsynthleads \abs{\iota}{x}{T}{T'}} &
    \infer{\Gamma\vdash \beta\{t'\} \tpcheck T}
          {T \leadsto_\beta^* \{t_1 \simeq t_2\} \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}
          {\Gamma\vdash t\tpcheck \{\absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}} &
    \infer{\Gamma \vdash \rho\ t\ @x\langle t_2 \rangle.T'\ \mhyph\ t' \tpcheck T}
          {
          \begin{array}{ccc}
            \Gamma \vdash t \tpsynthleads \{t_1 \simeq t_2'\}
            & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
            & |t_2'| =_{\beta\eta} |t_2|
            \\ \Gamma \vdash [t_2/x] T' \tpsynth \star
            & \Gamma \vdash t' \tpcheck [t_2/x]T'
            & [t_1/x]T' \cong T
          \end{array}
          } \\ \\
    \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
          {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T} &
    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T}
          {
              \Gamma\vdash t\tpcheck \{t' \simeq t''\}
              & \Gamma\vdash t' \Leftrightarrow T
              & \textit{FV}(t'') \subseteq \textit{dom}(\Gamma)
          }
  \end{array}
    \end{array}
  \]
\caption{Rules for checking a term against a type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

@@

\begin{figure}
  \[
    \begin{array}{c}
      \fbox{\(T_1 \cong T_2\)}
      \quad \fbox{\(T_1 \cong^{\text{t}} T_2\)}
      \\ \\
      \infer{
       T_1 \cong T_2
      }{
       T_1 \leadsto^*_\beta T_1' \nleadsto_\beta
       \quad T_2 \leadsto^*_\beta T_2' \nleadsto_\beta
       \quad T_1' \cong^{\text{t}} T_2'
      }
      \\ \\
      \begin{array}{cc}
        \infer{X \cong^{\text{t}} X}{}
        &
        \infer{
         \abs{\forall}{X}{\kappa_1}{T_1} \cong^{\text{t}} \abs{\forall}{X}{\kappa_2}{T_2}
        }{
         \kappa_1 \cong \kappa_2
         \quad  T_1 \cong T_2
        }
        \\ \\
          \infer{
           \abs{\forall}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\forall}{x}{T_2}{T_2'}
          }{
           T_1 \cong T_2
           \quad  T_1' \cong T_2'
          }
        & 
        \infer{
         \abs{\Pi}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\Pi}{x}{T_2}{T_2'}
        }{
         T_1 \cong T_2
         \quad  T_1' \cong T_2'
        }
        \\ \\ 
          \infer{
           \abs{\lambda}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\lambda}{x}{T_2}{T_2'}
          }{
           T_1 \cong T_2
           \quad  T_1' \cong T_2'
          }
        & 
        \infer{
         \abs{\lambda}{X}{\kappa_1}{T_1} \cong^{\text{t}} \abs{\lambda}{X}{\kappa_2}{T_2}
        }{
         \kappa_1 \cong \kappa_2
         \quad  T_1 \cong T_2
        }
        \\ \\
          \infer{
          \abs{\iota}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\iota}{x}{T_2}{T_2'}
          }{
          T_1 \cong T_2
          \quad  T_1' \cong T_2'
          }
        & 
        \infer{
         T_1\ t_1 \cong^{\text{t}} T_2\ t_2
        }{
         T_1 \cong^{\text{t}} T_2 \quad |t_1| =_{\beta\eta} |t_2|
        }
        \\ \\
          \infer{
           T_1 \cdot T_1' \cong^{\text{t}} T_2 \cdot T_2'
          }{
           T_1 \cong^{\text{t}} T_2
           \quad T_1' \cong T_2'
          }
        &
        \infer{
         \{ t_1 \simeq t_2 \} \cong^{\text{t}} \{ t_1'\ \simeq t_2' \}
        }{
         |t_1| =_{\beta\eta} |t_1'| \quad |t_2| =_{\beta\eta} |t_2'|
        }
      \end{array}
      \\ \\ \fbox{\(\kappa_1 \cong \kappa_2\)}
      \\ \\
      \begin{array}{cc}
        \infer{
         \star \cong \star
        }{}
        & \infer{
           \abs{\Pi}{x}{T_1}{\kappa_1} \cong \abs{\Pi}{x}{T_2}{\kappa_2}
          }{
           T_1 \cong T_2
           \quad \kappa_1 \cong \kappa_2
          }
        \\ \\
        \infer{
         \abs{\Pi}{X}{\kappa_1}{\kappa_1'} \cong \abs{\Pi}{X}{\kappa_2}{\kappa_2'}
        }{
         \kappa_1 \cong \kappa_2
         \quad \kappa_1' \cong \kappa_2'
        }
      \end{array}
    \end{array}
  \]
  \caption{Conversion rules for classifiers}
  \label{fig:conv}
\end{figure}

@@

Figure~\ref{fig:semtp} gives a realizability semantics for types and
kinds, following the semantics given in the previous papers on
CDLE~\cite{stump18,stump17}.  Details of this semantics are presented
further in Section~\ref{sec:snd} below.  Using the semantics and the
definition in Figure~\ref{fig:semctxt} of $\interp{\Gamma}$, we can
prove the following theorem:
\begin{theorem}[Soundness]
\label{thm:snd}
Suppose $(\sigma,\rho)\in\interp{\Gamma}$.  Then we have:
\begin{enumerate}
\item If $\Gamma\vdash \kappa$, then $\interp{\kappa}_{\sigma,\rho}$ is defined.
\item If $\Gamma\vdash T \tpsynth \kappa$, then $\interp{T}_{\sigma,\rho}\in\interp{\kappa}_{\sigma,\rho}$.
\item If $\Gamma\vdash t \tpsynth T$ then $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}\in \mathcal{R}$.
\item If $\Gamma\vdash t \tpcheck T$ and $\interp{T}_{\sigma,\rho}\in \mathcal{R}$, then
    $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}\in \mathcal{R}$.
\item If $T \cong T'$ or $T \cong^{\text{t}} T'$ and $\interp{T}_{\sigma,\rho}$ and $\interp{T'}_{\sigma,\rho}$ are both defined, then they are equal.
\item If \(\kappa \cong \kappa'\) and \(\interp{\kappa}_{\sigma,\rho}\) and
  \(\interp{\kappa'}_{\sigma,\rho}\) are both defined, then they are equal.
\end{enumerate}
\end{theorem}

@@

\begin{theorem}[Syntactic kind preservation]
  \label{thm:syntactic-kind-pres}
  If \(\Gamma \vdash T \tpsynth \kappa\) and \(T \leadsto_{\beta} T'\) then
  \(\Gamma \vdash T' \tpsynth \kappa'\) for some \(\kappa'\) such that \(\kappa
  \cong \kappa'\).
\end{theorem}

@@

\begin{lemma}
  \label{lem:interppres}
  If $T \leadsto^*_{\beta} T'$ and $\interp{T}_{\sigma,\rho}$ is defined, then $\interp{T'}_{\sigma,\rho}$ is also defined and equals $\interp{T}_{\sigma,\rho}$.
\end{lemma}
\begin{proof}
  This follows by induction on the reduction derivation, making use of the previous substitution lemmas.
\end{proof}

@@

\startcase{.2cm}
\[
  \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck T}{T \leadsto_\beta^* \abs{\Pi}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t\tpcheck T_2}
\]
To show $[\sigma \lambda
x.|t|]_{\cbe}\in\interp{\abs{\Pi}{x}{T_1}{T_2}}_{\sigma,\rho}$ (noting that the
latter is defined and in $\mathcal{R}$ by assumption), it suffices to assume an
arbitrary $E\in\interp{T_1}_{\sigma,\rho}$, and show
$[[\choice(E)/x]\sigma|t|]_{\cbe}\in\interp{T_2}_{\sigma[x\mapsto\choice(E)],\rho}$.
By the IH, we have
$[\sigma[x\mapsto\choice(E)]|t|]_{\cbe}\in\interp{T_2}_{\sigma[x\mapsto\choice(E)],\rho}$.
But $[\sigma[x\mapsto\choice(E)]t]_{\cbe} = [[\choice(E)/x]\sigma|t|]_{\cbe}$,
so this is sufficient.

@@

\startcase{.2cm}
\[
  \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck T'}
  {T' \leadsto_\beta^* \abs{\forall}{X}{\kappa}{T} & \Gamma,X:\kappa\vdash t \tpcheck T}
\]
By the IH, $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho[X\mapsto S]}$, for all $S\in\interp{\kappa}_{\sigma,\rho}$.
This is sufficient to prove $[\sigma |\absu{\Lambda}{X}{t}|]_{\cbe}\in\interp{\forall X:\kappa.T}_{\sigma,\rho}$, by the semantics
of $\forall$-types and definition of erasure.

@@

\startcase{.2cm}
\[
  \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck T}
  {T \leadsto_\beta^* \abs{\forall}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t \tpcheck T_2 & x\not\in\textit{FV}(|t|)}
\]

@@

\startcase{.2cm}
\[
  \infer{\Gamma\vdash [ t_1 , t_2 ] \tpcheck T}
  {
    \begin{array}{cc}
      T \leadsto_\beta^* \abs{\iota}{x}{T_1}{T_1}
      & \Gamma \vdash t_1 \tpcheck T_1
      \\ \Gamma \vdash t_2 \tpcheck [t/x]^{T_1}\ T_2
      & |t_1| =_{\beta\eta} |t_2|
    \end{array}
  }
\]
By the IH, we have $[\sigma |t_1|]_{\cbe}\in\interp{T_1}_{\sigma,\rho}$ and
$[\sigma |t_2|]_{\cbe}\in\interp{[t_1/x]T_2}_{\sigma,\rho}$.
By Lemma~\ref{lem:termsubstinterp}, the latter is equivalent to
$[\sigma |t|]_{\cbe}\in\interp{T_2}_{\sigma[x\mapsto\choice([\sigma
  |t|]_{\cbe}),\rho}$.
These two facts about $[\sigma |t|]_{\cbe}$ are
sufficient, by the semantics of $\iota$-types, for the desired
conclusion, using also the fact (from the fourth premise) that $\sigma|t| =_{\cbe} \sigma|t'|$.

@@

\startcase{.2cm}
\[
  \infer{\Gamma\vdash \beta\{t'\} \tpcheck T}
  {T \leadsto_\beta^* \{t_1 \simeq t_2\} \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}
\]
By the third premise \(|t_1| =_{\beta\eta} |t_2|\) it follows that \(\sigma|t_1|
=_{\beta\eta} \sigma|t_2|\), and also by the second premise and the fact that
\(\sigma \in \interp{\Gamma}\) it follows that
\(\textit{FV}(t_1\ t_2) \subseteq \textit{dom}(\sigma)\).
Also, we see \(\sigma|t'|\) is a closed term.
So, $[\sigma|t'|]_{\cbe}\in\interp{\{ t_1 \simeq t_2 \}}_{\sigma,\rho}$ follows
directly from the semantics of equality types.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpsynth \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  
\]
By the semantics of equality types, $[\sigma|t|]_{\cbe}$ cannot be in the interpretation of the equation in the premise,
since the two terms in question are closed and not $\beta\eta$-equal.  By the IH applied to the first premise, however,
$[\sigma|t|]_{\cbe}$ is in the interpretation of that equation.  This is a contradiction.

>>>>

\newcommand{\choice}[0]{\zeta}
\newcommand{\elcap}[0]{\cap}
\newcommand{\abs}[4]{{#1}\, #2\! : \! #3.\, #4}
\newcommand{\absu}[3]{{#1}\, #2.\, #3}
\mathchardef\mhyph="2D % Define a "math hyphen"
\newcommand{\interp}[1]{\llbracket #1 \rrbracket} 
\newcommand{\leadstoc}[0]{\ensuremath{\leadsto_{\mathbf{n}}}}
\newcommand{\leadstocs}[0]{\ensuremath{\leadsto_{\mathbf{n}}^*}}
\newcommand{\nleadstoc}[0]{\ensuremath{\nleadsto_{\mathbf{n}}}}
\newcommand{\tpcheck}[0]{\Leftarrow}
\newcommand{\tpsynth}[0]{\Rightarrow}
\newcommand{\tpsynthleads}[0]{\ensuremath{\overset{\leadstocs}{\Rightarrow}}}
\newcommand{\cbe}[0]{c\beta\eta}
\newcommand{\utp}[0]{\mathcal{U}}
\newcommand{\startcase}[1]{\vspace{#1} \noindent\textbf{\underline{Case:}}}
\newcommand{\startsubcase}[1]{\vspace{#1} \noindent\textbf{\underline{Subcase:}}}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{observation}[theorem]{Observation}

@@

In terms, when the type of an introduction form is checked or the type of an
elimination form is synthesized, we use call-by-name normalization, written
\(\leadstoc\) and \(\leadstocs\) for the reflexive transitive closure, to put
types in weak head normal form, revealing type constructors.
We abbreviate the conjuction ``term \(t\) synthesizes some type, and that type
call-by-name reduces to another type'' with the symbol \(\tpsynthleads\),
defined formally at the top of Figure~\ref{fig:tp}.
When a redex (reducible expression) occurs such that the argument to a type is a
term, such as \((\abs{\lambda}{x}{T_1}{T_2})\ t_1\), the reduction uses
substitution with annotations to \([t_1/x]^{T_1}\ T_2\).

@@

\begin{figure}
  \[
    \begin{array}{c}
      \Gamma \vdash t \tpsynthleads T\ =\ \exists T'.\ \Gamma
      \vdash t \tpsynth T' \land T' \leadstocs T
      \\ \\ \\
  \begin{array}{cc}
    \infer{\Gamma\vdash x\tpsynth T}{(x : T)\in\Gamma} &
    \infer{\Gamma\vdash t\tpcheck T}{\Gamma\vdash t\tpsynth T' & T' \cong T} \\ \\
    \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck T}{T \leadstocs \abs{\Pi}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t\tpcheck T_2} &
    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynthleads \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\

    \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck T'}
          {T' \leadstocs \abs{\forall}{X}{\kappa}{T} & \Gamma,X:\kappa\vdash t \tpcheck T} &
    \infer{\Gamma\vdash t \cdot T' \tpsynth [T'/X]T}
          {\Gamma\vdash t \tpsynthleads \abs{\forall}{X}{\kappa}{T} & \Gamma\vdash T' \tpsynth \kappa' & \kappa'\cong\kappa} \\ \\

    \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck T}
          {T \leadstocs \abs{\forall}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t \tpcheck T_2 & x\not\in\textit{FV}(|t|)} &
    \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynthleads \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\

    \infer{\Gamma\vdash [ t_1 , t_2 ] \tpcheck T}
          {
           \begin{array}{cc}
             T \leadstocs \abs{\iota}{x}{T_1}{T_1}
             & \Gamma \vdash t_1 \tpcheck T_1
             \\ \Gamma \vdash t_2 \tpcheck [t_1/x]^{T_1}\ T_2
             & |t_1| =_{\beta\eta} |t_2|
           \end{array}
          } &
    \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynthleads \abs{\iota}{x}{T}{T'}} \\ \\
    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x] T'}{\Gamma\vdash t \tpsynthleads \abs{\iota}{x}{T}{T'}} &
    \infer{\Gamma\vdash \beta\{t'\} \tpcheck T}
          {T \leadstocs \{t_1 \simeq t_2\} \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}
          {
            \Gamma \vdash t \tpsynth T'
            \quad T' \cong \{\absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}
          } &
    \infer{\Gamma \vdash \rho\ t\ @x\langle t_2 \rangle.T'\ \mhyph\ t' \tpcheck T}
          {
          \begin{array}{ccc}
            \Gamma \vdash t \tpsynthleads \{t_1 \simeq t_2'\}
            & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
            & |t_2'| =_{\beta\eta} |t_2|
            \\ \Gamma \vdash [t_2/x] T' \tpsynth \star
            & \Gamma \vdash t' \tpcheck [t_2/x]T'
            & [t_1/x]T' \cong T
          \end{array}
          } \\ \\
    \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
          {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T} &
    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T}
          {
              \Gamma\vdash t\tpcheck \{t' \simeq t''\}
              & \Gamma\vdash t' \Leftrightarrow T
              & \textit{FV}(t'') \subseteq \textit{dom}(\Gamma)
          }
  \end{array}
    \end{array}
  \]
\caption{Rules for checking a term against a type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

@@

\begin{figure}
  \[
    \begin{array}{c}
      \fbox{\(T_1 \cong T_2\)}
      \quad \fbox{\(T_1 \cong^{\text{t}} T_2\)}
      \\ \\
      \infer{
       T_1 \cong T_2
      }{
       T_1 \leadstocs T_1' \nleadstoc
       \quad T_2 \leadstocs T_2' \nleadstoc
       \quad T_1' \cong^{\text{t}} T_2'
      }
      \\ \\
      \begin{array}{cc}
        \infer{X \cong^{\text{t}} X}{}
        &
        \infer{
         \abs{\forall}{X}{\kappa_1}{T_1} \cong^{\text{t}} \abs{\forall}{X}{\kappa_2}{T_2}
        }{
         \kappa_1 \cong \kappa_2
         \quad  T_1 \cong T_2
        }
        \\ \\
          \infer{
           \abs{\forall}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\forall}{x}{T_2}{T_2'}
          }{
           T_1 \cong T_2
           \quad  T_1' \cong T_2'
          }
        & 
        \infer{
         \abs{\Pi}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\Pi}{x}{T_2}{T_2'}
        }{
         T_1 \cong T_2
         \quad  T_1' \cong T_2'
        }
        \\ \\ 
          \infer{
           \abs{\lambda}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\lambda}{x}{T_2}{T_2'}
          }{
           T_1 \cong T_2
           \quad  T_1' \cong T_2'
          }
        & 
        \infer{
         \abs{\lambda}{X}{\kappa_1}{T_1} \cong^{\text{t}} \abs{\lambda}{X}{\kappa_2}{T_2}
        }{
         \kappa_1 \cong \kappa_2
         \quad  T_1 \cong T_2
        }
        \\ \\
          \infer{
          \abs{\iota}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\iota}{x}{T_2}{T_2'}
          }{
          T_1 \cong T_2
          \quad  T_1' \cong T_2'
          }
        & 
        \infer{
         T_1\ t_1 \cong^{\text{t}} T_2\ t_2
        }{
         T_1 \cong^{\text{t}} T_2 \quad |t_1| =_{\beta\eta} |t_2|
        }
        \\ \\
          \infer{
           T_1 \cdot T_1' \cong^{\text{t}} T_2 \cdot T_2'
          }{
           T_1 \cong^{\text{t}} T_2
           \quad T_1' \cong T_2'
          }
        &
        \infer{
         \{ t_1 \simeq t_2 \} \cong^{\text{t}} \{ t_1'\ \simeq t_2' \}
        }{
         |t_1| =_{\beta\eta} |t_1'| \quad |t_2| =_{\beta\eta} |t_2'|
        }
      \end{array}
      \\ \\ \fbox{\(\kappa_1 \cong \kappa_2\)}
      \\ \\
      \begin{array}{cc}
        \infer{
         \star \cong \star
        }{}
        & \infer{
           \abs{\Pi}{x}{T_1}{\kappa_1} \cong \abs{\Pi}{x}{T_2}{\kappa_2}
          }{
           T_1 \cong T_2
           \quad \kappa_1 \cong \kappa_2
          }
        \\ \\
        \infer{
         \abs{\Pi}{X}{\kappa_1}{\kappa_1'} \cong \abs{\Pi}{X}{\kappa_2}{\kappa_2'}
        }{
         \kappa_1 \cong \kappa_2
         \quad \kappa_1' \cong \kappa_2'
        }
      \end{array}
    \end{array}
  \]
  \caption{Conversion rules for classifiers}
  \label{fig:conv}
\end{figure}

@@

Figure~\ref{fig:semtp} gives a realizability semantics for types and
kinds, following the semantics given in the previous papers on
CDLE~\cite{stump18,stump17}.  Details of this semantics are presented
further in Section~\ref{sec:snd} below.  Using the semantics and the
definition in Figure~\ref{fig:semctxt} of $\interp{\Gamma}$, we can
prove the following theorem:
\begin{theorem}[Soundness]
\label{thm:snd}
Suppose $(\sigma,\rho)\in\interp{\Gamma}$.  Then we have:
\begin{enumerate}
\item If $\Gamma\vdash \kappa$, then $\interp{\kappa}_{\sigma,\rho}$ is defined.
\item If $\Gamma\vdash T \tpsynth \kappa$, then $\interp{T}_{\sigma,\rho}\in\interp{\kappa}_{\sigma,\rho}$.
\item If $\Gamma\vdash t \tpsynth T$ then $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}\in \mathcal{R}$.
\item If $\Gamma\vdash t \tpcheck T$ and $\interp{T}_{\sigma,\rho}\in \mathcal{R}$, then
    $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}$.
\item If $T \cong T'$ or $T \cong^{\text{t}} T'$ and $\interp{T}_{\sigma,\rho}$ and $\interp{T'}_{\sigma,\rho}$ are both defined, then they are equal.
\item If \(\kappa \cong \kappa'\) and \(\interp{\kappa}_{\sigma,\rho}\) and
  \(\interp{\kappa'}_{\sigma,\rho}\) are both defined, then they are equal.
\end{enumerate}
\end{theorem}

@@

\begin{theorem}[Syntactic kind preservation]
  \label{thm:syntactic-kind-pres}
  If \(\Gamma \vdash T \tpsynth \kappa\) and \(T \leadstoc T'\) then
  \(\Gamma \vdash T' \tpsynth \kappa'\) for some \(\kappa'\) such that \(\kappa
  \cong \kappa'\).
\end{theorem}

@@

\begin{lemma}
  \label{lem:interppres}
  If $T \leadstocs T'$ and $\interp{T}_{\sigma,\rho}$ is defined, then $\interp{T'}_{\sigma,\rho}$ is also defined and equals $\interp{T}_{\sigma,\rho}$.
\end{lemma}
\begin{proof}
  This follows by induction on the reduction derivation, making use of the previous substitution lemmas.
\end{proof}

@@

\startcase{.2cm}
\[
  \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck T}{T \leadstocs \abs{\Pi}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t\tpcheck T_2}
\]
To show $[\sigma \lambda
x.|t|]_{\cbe}\in\interp{\abs{\Pi}{x}{T_1}{T_2}}_{\sigma,\rho}$ (noting that the
latter is defined and in $\mathcal{R}$ by assumption), it suffices to assume an
arbitrary $E\in\interp{T_1}_{\sigma,\rho}$, and show
$[[\choice(E)/x]\sigma|t|]_{\cbe}\in\interp{T_2}_{\sigma[x\mapsto\choice(E)],\rho}$.
By the IH, we have
$[\sigma[x\mapsto\choice(E)]|t|]_{\cbe}\in\interp{T_2}_{\sigma[x\mapsto\choice(E)],\rho}$.
But $[\sigma[x\mapsto\choice(E)]t]_{\cbe} = [[\choice(E)/x]\sigma|t|]_{\cbe}$,
so this is sufficient.

@@

\startcase{.2cm}
\[
  \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck T'}
  {T' \leadstocs \abs{\forall}{X}{\kappa}{T} & \Gamma,X:\kappa\vdash t \tpcheck T}
\]
By the IH, $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho[X\mapsto S]}$, for all $S\in\interp{\kappa}_{\sigma,\rho}$.
This is sufficient to prove $[\sigma |\absu{\Lambda}{X}{t}|]_{\cbe}\in\interp{\forall X:\kappa.T}_{\sigma,\rho}$, by the semantics
of $\forall$-types and definition of erasure.

@@

\startcase{.2cm}
\[
  \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck T}
  {T \leadstocs \abs{\forall}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t \tpcheck T_2 & x\not\in\textit{FV}(|t|)}
\]

@@

\startcase{.2cm}
\[
  \infer{\Gamma\vdash [ t_1 , t_2 ] \tpcheck T}
  {
    \begin{array}{cc}
      T \leadstocs \abs{\iota}{x}{T_1}{T_1}
      & \Gamma \vdash t_1 \tpcheck T_1
      \\ \Gamma \vdash t_2 \tpcheck [t/x]^{T_1}\ T_2
      & |t_1| =_{\beta\eta} |t_2|
    \end{array}
  }
\]
By the IH, we have $[\sigma |t_1|]_{\cbe}\in\interp{T_1}_{\sigma,\rho}$ and
$[\sigma |t_2|]_{\cbe}\in\interp{[t_1/x]T_2}_{\sigma,\rho}$.
By Lemma~\ref{lem:termsubstinterp}, the latter is equivalent to
$[\sigma |t|]_{\cbe}\in\interp{T_2}_{\sigma[x\mapsto\choice([\sigma
  |t|]_{\cbe}),\rho}$.
These two facts about $[\sigma |t|]_{\cbe}$ are
sufficient, by the semantics of $\iota$-types, for the desired
conclusion, using also the fact (from the fourth premise) that $\sigma|t| =_{\cbe} \sigma|t'|$.

@@

\startcase{.2cm}
\[
  \infer{\Gamma\vdash \beta\{t'\} \tpcheck T}
  {T \leadstocs \{t_1 \simeq t_2\} \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}
\]
By the third premise \(|t_1| =_{\beta\eta} |t_2|\) it follows that \(\sigma|t_1|
=_{\beta\eta} \sigma|t_2|\), and also by the second premise and the fact that
\(\sigma \in \interp{\Gamma}\) it follows that
\(\textit{FV}(t_1\ t_2) \subseteq \textit{dom}(\sigma)\).
Also, we see \(\sigma|t'|\) is a closed term.
So, $[\sigma|t'|]_{\cbe}\in\interp{\{ t_1 \simeq t_2 \}}_{\sigma,\rho}$ follows
directly from the semantics of equality types.
